### 2025-04-04 latex/HP3_comm.tex
new files

%%%%%%%%%%
\subsection*{3.3}
\begin{translation}[hp03_003]
[\dots] the empty pathway becomes the royal highway for \emph{prāṇa}, the mind becomes free of support, and death is cheated.
\end{translation}

@@

%\begin{philcomm}[hp03_007]
%JB: hi (in 3.7c) is supported only by the \textit{Jyotsnā}. Mss. have tu and ca.
%\end{philcomm}

@@

%%%%%%%%%%
\subsection*{3.10 heading}
\begin{translation}[hp03_010a]
Now the great seal (\emph{mahāmudrā}):
\end{translation}

@@

\begin{philcomm}[hp03_012]
This verse is taken from the \emph{Amaraugha}, which uses the alchemical imagery of the \emph{Amṛtasiddhi} to describe the stilling of Kuṇḍalinī in the central channel. Drawing on Hellwig 2009:238–240, Mallinson and Szanto (2021:21) note that “In alchemical texts \emph{māraṇa} (“killing”) involves heating a substance and thereby changing its state, usually through calcination or oxidation, so that it becomes inert. In the \emph{Amṛtasiddhi māraṇa} and other derivatives of the root \emph{mṛ}, “die”, are used to denote the stilling or stopping of either the breath or Bindu.” Thus when Kuṇḍalinī is said to be in the state of \emph{maraṇa} the meaning is that she is stilled. The \emph{dvipuṭa} or “vessel with two halves” in which this occurs is the same as the \emph{Amṛtasiddhi}’s \emph{saṃpuṭa}, which, drawing on Hellwig 2009:342, Mallinson and Szanto (2021:22) say “consists of two \emph{puṭa}s joined together to form a sealed crucible for heating reagents without evaporation”. In the yoga of the \emph{Amṛtasiddhi}, the bodily \emph{saṃpuṭa} is formed by applying locks at the top and bottom of the central channel, i.e.~constricting the perineal region and the throat. In the \emph{Haṭhapradīpikā} it is formed by pressing the perineum with the heel and constricting the throat.

As Birch (2019: 971) notes, it is unlikely that later non-Buddhist authors understood \textit{maraṇāvasthā} and \emph{dvipuṭa} according to the alchemical metaphors of the \textit{Amṛtasiddhi}. Later commentators take \emph{dvipuṭa} as the two nostrils (e.g., \textit{Yuktabhavadeva} 7.187, \textit{dvināsāpuṭa}); the \textit{iḍā} and \textit{piṅgalā} channels (e.g., \textit{Jyotsnā} 3.27, \textit{puṭayor dvayam iḍāpiṅgalayor yugmam}); or the in and out flows of the breath (e.g., \textit{Yogaprakāśikā} 5.16–17, \textit{vāyor bahirnirgamanam antaḥpraveśa iti yat puṭadvayaṃ tam}). How these commentators understood \textit{asau māraṇāvasthā} is less clear. Brahmānanda seems to take it as the death of \textit{prāṇa}, or in other words, the absence of the breath, in the two nostrils (\textit{maraṇāvasthā jāyate kuṇḍalībodhe sati suṣumnāyāṃ praviṣṭe prāṇe dvayoḥ puṭayoḥ prāṇaviyogāt}).  Bhavadeva thought that Kuṇḍalinī, along with \textit{prāṇa} and \textit{apāna}, remains in the two nostrils while the breath is being held (\textit{evaṃ vāyudhāraṇāyāṃ kriyamāṇāyāṃ vyākulā bhūtā Kuṇḍalinī apānaprāṇābhyāṃ saha nāsāpuṭadvayāśritā bhavati}). Others, such as Śivānanda and Bālakṛṣṇa, favour the reading \textit{tadā sā maraṇāvasthāṃ harate dvi\-puṭāśritām} (or \textit{dvipuṭāśrayām}), which is present in group 2 and 3 manuscripts of the \textit{Haṭhapradīpikā}. Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened Kuṇḍalinī destroys death, which is usually dependent on the in and out breaths.

%JB: I think our translation should reflect a likely Śaiva interpretation rather than the alchemical metaphors in the Amṛtasiddhi (e.g., `Then, that state of death arises in the two nostrils'). JM: this would imply that we think that Svātmārāma decided to use the verse with this understanding, which seems unlikely to me. JB. Unlikely? All the evidence in commentaries on the HP and later compendiums, like the YCM and Yuktabhavadeva, indicates that the alchemical metaphors of the AS were not understood here. It's not even clear that the alchemical metaphors work with raising moving or raising kuṇḍalinī. In contrast to this, there's not one commentarial explanation or rewrite of the verse that reveals an alchemical meaning. We also have to bear in mind that there's no evidence that Svātmārāma and later commentators knew the AS. In the one case where there is some evidence (i.e., Śivānanda quotes it in the YCM), it doesn't seem to have helped the author understand the alchemical metaphor in this instance (as Śivānanda rewrote the verse to say that kuṇḍalinī destroys the state of death).

@@

%%%%%%%%%%
\subsection*{3.14}
\begin{translation}[hp03_014]
Problems such as the great afflictions [and] death and so forth are removed [by it], and that is why the most wise call it the `great seal.'
\end{translation}

@@

\begin{philcomm}[hp03_018]
Manuscripts of three groups (\textalpha, \textbeta\ and \textgamma), which are important stemmatically, have the reading \emph{jarāmṛtyuvināśinī} (and \getsiglum{J5} has \emph{nṛṇāṃ mṛtyuvināśinī}) for the second verse quarter. While this reading is possible, the play on \emph{mahāsiddhi} and \emph{mahāmudrā} seems more likely original, as seen in the source text, the \emph{Vivekamārtaṇḍa} (without significant variants), and the \texteta\ and \textepsilon\ groups.

@@

\begin{philcomm}[hp03_020]
The referent of \emph{madhye} is uncertain. The verse is derived from the \emph{Amaraugha}, and the \emph{Amṛtasiddhi} makes no mention of a place to focus the mind in its treatment of \emph{mahābandha} (it does however instruct the yogi to place the mind at the \emph{catuṣpatha} in its teachings on \emph{mahāmudrā}). Bhavadevamiśra (7.196), Brahmānanda (3.20) and Bālakṛṣṇa (5.24) take it to mean the central channel. It could also plausibly mean the region between the chest and perineum, or perhaps the place between the eyebrows. At 3.24 this practice is said to make the mind reach Kedāra, which is sometimes located between the eyebrows (see Mallinson 2007:214 n.285, Birch 2019: 967 n.57).
\end{philcomm}

%%%%%%%%%%
\subsection*{3.21}
\begin{translation}[hp03_021]
[The yogi] should hold the breath as long as possible and exhale slowly. And having practised it on the left side, he should practise it on the right side. 
\end{translation}

@@

\begin{philcomm}[hp03_021]
This section in the description of \emph{mahābandha} has been subject to various revisions in the manuscript groups. The shortest version occurs in \alphaOne\ and \alphaTwo, as well as the \textbeta\ group and the \emph{Haṭharatnāvalī}, which omit verses 3.21 and 3.22. This omission results in the somewhat odd juxtaposition of two lines stating that this is \emph{mahābandha}.

@@

\begin{philcomm}[hp03_022]
This verse expresses an alternative to the application of the chin lock in \emph{mahā\-bandha} (mentioned in 3.20). It is found in all groups but is notably absent in \alphaOne\ and \alphaTwo, and the \textbeta\ group. At present we cannot be certain if this verse was written by Svātmārāma because it does not reflect textual teachings on \emph{mahābandha} in so far as no other text known to us advocates the use of the tongue rather than the chin lock in \emph{mahābandha}. If the verse was composed by Svātmārāma, it was omitted early in the transmission by someone who did not agree with the alternative teaching.

This verse occurs in the \emph{Yogacintāmaṇi} in the middle of a quotation attributed to Īśvara. The other verses of the quoted passage are found in the \emph{Śivasaṃhitā} (4.37–42), but the verse in question is not reported in the critical edition of the \emph{Śivasaṃhitā} (2018). The verse is absent in another passage on \emph{mahābandha} that the author of the \emph{Yogacintāmaṇi} cites and attributes to the \emph{Haṭhapradīpikā}.

@@

\begin{philcomm}[hp03_024]
The reading \emph{ūrdhvaṃgatibodhakaḥ} (`initiates an upward flow') is found in the source (\emph{Amaraugha} 25) and all \emph{Haṭhapradīpikā} witnesses except the \emph{Jyotsnā} (where the line is found earlier). While the \emph{Amaraughaprabodha} has the reading \emph{ūrdhvaṃgativiśodhanaḥ} (`purification of the upward flow'), the \emph{Amṛtasiddhi} (12.14) states that the chin-lock prevents the upward flow (\emph{ūrdhvaṃgatinirodhakaḥ}) in all the channels. The \textit{Amaraugha} is referring to the idea (likely accepted by Svātmārāma) that the root lock creates an upward flow in all the channels that prevents the elements and essences of the body from escaping (cf.~\emph{Amṛtasiddhi} 12.8–10).

Triveṇī and Kedāra are pilgrimage sites, the former at Prayāga where the Gaṅgā and Yamunā meet, the latter in the Himālaya, near the source of the Gaṅgā. The bodily \emph{triveṇī} is located in the navel or heart by earlier Śaiva works (Birch 2019: 967). Here it may be the same as the \emph{trikūṭa} and located between the eyebrows (Mallinson 2007: 209 n.259). Brahmānanda, who does not identify a location for Triveṇī, understands Kedāra to be between the eyebrows. In the \emph{Khecarīvidyā} it is located on the back of the head above the nape of the neck (Mallinson 2007:214 n.285). For other references on the location of Kedāra, see Birch 2019:967 n.57.
\end{philcomm}

@@

\begin{philcomm}[hp03_027]
The term \emph{puṭadvaya} is referring back to the \emph{dvipuṭa} mentioned in verse 3.12. On the alchemical meaning and other interpretations by later commentators, see the note to 3.12.

@@

\begin{philcomm}[hp03_028]
The \textalpha\ and \textgamma\ groups have °\emph{saṃbandhāj}, which is possible but somewhat awkward (i.e., `because of the union of the moon, sun and fire, the state of a dead person, which has arisen, leads to immortality'). The adopted reading °\emph{saṃbandho}, which is supported by \epsilonTwo, \emph{Yuktabhavadeva} 7.200 and \emph{Jyotsna} 3.28, makes better sense but its meaning is not as clear as the formulation in the source text (i.e., \emph{Amaraugha} 30).

The compound \emph{mṛtāvasthā} (`the state of death') likely refers to a dead person in the sense that the yogi appears as if dead when the moon, sun and fire have united. In the version found in \emph{Jyotsnā} 3.28 and \etaOne, \etaTwo\ and \epsilonThree, the final verse quarter has been rewritten to say that the yogi then exhales the breath (\emph{tato vāyuṃ virecayet}). This implies that \emph{mṛtāvasthā} is a breath retention (\emph{kumhaka}), which is apparent in Brahmānanda's explanation:
\begin{quote}
The state of death that has arisen is the state of one who has died, [that is,] of one who is devoid of the breath because of the absence of movement of \emph{prāṇa} in the \emph{iḍā} and \emph{piṅgalā} channels. Then, immediately after that [state], [the yogi] exhales the breath, [that is,] he gradually releases it through the nostrils.\\
\emph{mṛtasya prāṇaviyuktasyāvasthā mṛtāvasthā samutpannā bhavati, iḍāpiṅgalayoḥ prāṇasañcārābhāvāt} | \emph{tatas tadanantaraṃ vāyuṃ virecayen nāsikāpuṭābhyāṃ śanais tyajet} |
\end{quote}

@@

%%%%%%%%%%
\subsection*{3.29}
\begin{translation}[hp03_029]
Through practice, this great piercing bestows the great \emph{siddhi} [and] cures wrinkles, grey hair and trembling. It is used by the best practitioners.
\end{translation}

@@

\begin{philcomm}[hp03_029]
In the third verse quarter, the reading \emph{°vedhaghnaḥ} (`cures wounds' ?) in \textalpha\ and the other groups, the exception being \textgamma\ (\emph{°vegaghnaḥ}), is strange as it does not seem related to the other two symptoms of old age (i.e., wrinkles and grey hair) that this \emph{mudrā} can cure. It is perhaps a misreading of \emph{°vepaghnaḥ}, which occurs in some manuscripts in the \textbeta\ (\getsiglum{J1}, \getsiglum{N12}), \textepsilon\ (\getsiglum{J14}) and \textdelta\ (\getsiglum{J3}, \getsiglum{N16}, \getsiglum{N18}) groups, as well as the \emph{Yogacintāmaṇi} and \emph{Jyotsnā}. The meaning of \emph{vepa}, `trembling,' is consistent with wrinkles and grey hair.

%The collocation \emph{valīpalitaroga} that we have adopted, found in \getsiglum{C8}, is quite common (e.g. \emph{Śāradātilaka} \emph{6.104, Ānandakanda} 1.7.13, .1.7.31).
%JB: I think we should adopt vepa on the strength of the marmasthāna entry (\item[valīpalitavepaghnaḥ] Ba5, J1, J3, J6, J14, N4a, N11, N12, N16, N18, ), and the fact that it's closer to vedha (alpha reading) than roga (which looks like an outlier in the marmasthāna data). 
% J1=6b, J3,J6,N16,N18=6a, J14=4c, N11 =Gr2, N12 =4b,
% MD: These mss are not very strong. Is it not possible to interpret °vedhaghnaḥ as "destroys by piercing"? This verse only has no source in the middle of the long citation from Amaraugha, is probably authorial. Svātmārāma might have wanted to use the word vedha of mahāvedha, similar to dhauti/dhāvanti.
\end{philcomm}

@@

Various versions of this verse have been transmitted in \emph{āryā}, \emph{gīti} and \emph{anuṣṭubh} metres. We have adopted a version close to \alphaTwo, which has a slight metrical fault:
\begin{quote}
 chedanacālanadohaiḥ kalāṃ krameṇa pravardhayet tāvat |\\
 sā yāvad bhrūmadhyaṃ viśati tadānīṃ khecarīsiddhiḥ ||
\end{quote}
% alpha2
% chedanacālanaṃ dohaiḥ kalā kramaṇa pravardhaye tāvat
% sā yāvad bhrumadhyaṃ visati tadānī khecarīsiddhiḥ
The emendation of \emph{tadānīṃ} to \emph{tadā} renders the verse an \emph{āryā}. The word \emph{krameṇa} is well attested by manuscripts of the \textalpha, \textbeta, \texteta\ and  \textepsilon\ groups, which all have unmetrical or corrupted versions. 
% krameṇa is well attested
% close to alpha 2
% damaged in alpha 1: [damaged]ḥ kalāḥ krameṇa varddayet | sā yāva[d] bhṛmadhyaṃ viśa tadānīṃ khecarīsiddhiḥ || but it looks like someone has unsuccessfully tried to make it anuṣṭubh. Perhaps, suggests that a longer metre was original.

@@

%%%%%%%%%%
\subsection*{3.34*1}
\begin{translation}[hp03_034_1]
If the yogi turns back the tongue and remains [like that] for half an instant, he is instantly freed from disease, death, old age and the like.
\end{translation}

\begin{sources}[hp03_034_1]
\emph{Śivasaṃhitā} 3.91
\begin{versinnote}
\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
\tl{kṣaṇena mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_034_1]
\emph{Yogacintāmaṇi} f.~74r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{kalāṃ parāṅmukhīṃ kṛtvā tripathe parivartayet |\\+}
\tl{sā bhavet khecarī mudrā vyomacakraṃ tad ucyate ||\\+}
\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
\tl{viṣayair mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp03_034_1]
This verse is absent in \alphaOne\ and was likely added to a block of verses that Svātmārāma borrowed from the \textit{Vivekamārtaṇḍa} (47–51). 3.34*1 is in \alphaTwo\ and \alphaThree\ (\alphaOne\ omits 2.35 and the greyscaled verses) and also the \textgamma\ and \texteta\ groups. Other manuscripts have an additional line that gives \emph{vyomacakra} as an alternative name for \emph{khecarīmudrā}. This alternative name does not occur in any of the source texts known to have been used by Svātmārāma, but it may have been inspired by the name \emph{nabhomudrā}, which is what the \emph{Vivekamārtaṇḍa} calls \emph{khecarīmudrā} (\emph{Vivekamārtaṇḍa} 40).   
\end{philcomm}

>>>>

%%%%%%%%%%
\subsection*{3.3}
\begin{translation}[hp03_003]
{[}\dots] the empty pathway becomes the royal highway for \emph{prāṇa}, the mind becomes free of support, and death is cheated.
\end{translation}

@@

%\begin{philcomm}[hp03_007]
The reading of \textalpha\, \emph{idaṃ mudrādi°}, is not attested by the other manuscript groups but it makes sense in so far as 'locks' (\emph{bandha}) and 'actions' (\emph{karaṇa}) figure among the ten techniques taught in this chapter. It is also consistent with the reference to \emph{mudrādi} in 1.55. However, most manuscripts have readings, such as \emph{idaṃ hi mudrādaśakam}, that refer to the techniques of this chapter as only \emph{mudrā}s, and this is consistent with 3.104.

%JB: hi (in 3.7c) is supported only by the \textit{Jyotsnā}. Mss. have tu and ca.
%\end{philcomm}

@@

%%%%%%%%%%
\subsection*{3.10 heading}
\begin{translation}[hp03_010a]
Of these, the great seal (\emph{mahāmudrā}) [is now taught]: 
\end{translation}

@@

\begin{philcomm}[hp03_012]
This verse is taken from the \emph{Amaraugha}, which uses the alchemical imagery of the \emph{Amṛtasiddhi} to describe the stilling of Kuṇḍalinī in the central channel. Drawing on Hellwig 2009: 238–240, Mallinson and Szanto (2021: 21) note that “In alchemical texts \emph{māraṇa} (“killing”) involves heating a substance and thereby changing its state, usually through calcination or oxidation, so that it becomes inert. In the \emph{Amṛtasiddhi māraṇa} and other derivatives of the root \emph{mṛ}, “die”, are used to denote the stilling or stopping of either the breath or Bindu.” Thus when Kuṇḍalinī is said to be in the state of \emph{maraṇa} the meaning is that she is stilled. The \emph{dvipuṭa} or “vessel with two halves” in which this occurs is the same as the \emph{Amṛtasiddhi}’s \emph{saṃpuṭa}, which, drawing on Hellwig 2009: 342, Mallinson and Szanto (2021: 22) say “consists of two \emph{puṭa}s joined together to form a sealed crucible for heating reagents without evaporation”. In the yoga of the \emph{Amṛtasiddhi}, the bodily \emph{saṃpuṭa} is formed by applying locks at the top and bottom of the central channel, i.e.~constricting the perineal region and the throat. In the \emph{Haṭhapradīpikā} it is formed by pressing the perineum with the heel and constricting the throat.\lb

As Birch (2019: 971) notes, it is unlikely that later non-Buddhist authors understood \textit{maraṇāvasthā} and \emph{dvipuṭa} according to the alchemical metaphors of the \textit{Amṛtasiddhi}. Later commentators take \emph{dvipuṭa} as the two nostrils (e.g., \textit{Yuktabhavadeva} 7.187, \textit{dvināsāpuṭa}); the \textit{iḍā} and \textit{piṅgalā} channels (e.g., \textit{Jyotsnā} 3.27, \textit{puṭayor dvayam iḍāpiṅgalayor yugmam}); or the in and out flows of the breath (e.g., \textit{Yogaprakāśikā} 5.16–17, \textit{vāyor bahirnirgamanam antaḥpraveśa iti yat puṭadvayaṃ tam}). How these commentators understood \textit{asau māraṇāvasthā} is less clear. Brahmānanda seems to take it as the death of \textit{prāṇa}, or in other words, the absence of the breath, in the two nostrils (\textit{maraṇāvasthā jāyate kuṇḍalībodhe sati suṣumnāyāṃ praviṣṭe prāṇe dvayoḥ puṭayoḥ prāṇaviyogāt}).  Bhavadeva thought that Kuṇḍalinī, along with \textit{prāṇa} and \textit{apāna}, remains in the two nostrils while the breath is being held (\textit{evaṃ vāyudhāraṇāyāṃ kriyamāṇāyāṃ vyākulā bhūtā Kuṇḍalinī apānaprāṇābhyāṃ saha nāsāpuṭadvayāśritā bhavati}). Others, such as Śivānanda and Bālakṛṣṇa, favour the reading \textit{tadā sā maraṇāvasthāṃ harate dvi\-puṭāśritām} (or \textit{dvipuṭāśrayām}), which is present in group 2 and 3 manuscripts of the \textit{Haṭhapradīpikā}. Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened Kuṇḍalinī destroys death, which is usually dependent on the in and out breaths.\lb

The form \textit{ṛjvībhūtā} is non-Paninian (it should be \textit{ṛjūbhūtā}) but all witnesses have \textit{ṛjvī} or variants thereof. \alphaOne\ has \textit{ṛjvī bhūtvā} which is correct morphologically, but it does not make sense with \textit{bhavet} in the next \textit{pāda} and may be an attempt at correction by the redactor.

%JB: I think our translation should reflect a likely Śaiva interpretation rather than the alchemical metaphors in the Amṛtasiddhi (e.g., `Then, that state of death arises in the two nostrils'). JM: this would imply that we think that Svātmārāma decided to use the verse with this understanding, which seems unlikely to me. JB. Unlikely? All the evidence in commentaries on the HP and later compendiums, like the YCM and Yuktabhavadeva, indicates that the alchemical metaphors of the AS were not understood here. It's not even clear that the alchemical metaphors work with raising moving or raising kuṇḍalinī. In contrast to this, there's not one commentarial explanation or rewrite of the verse that reveals an alchemical meaning. We also have to bear in mind that there's no evidence that Svātmārāma and later commentators knew the AS. In the one case where there is some evidence (i.e., Śivānanda quotes it in the YCM), it doesn't seem to have helped the author understand the alchemical metaphor in this instance (as Śivānanda rewrote the verse to say that kuṇḍalinī destroys the state of death).JM:but the commentators etc. are all a lot later than the HP. I think it's better to give Svātmārāma the benefit of the doubt. I don't think he would have used the verse if he didn't understand it correctly — as you (JB) point out in the note the later attempts to understand maraṇa in a non-alchemical way are unconvincing.

@@

%%%%%%%%%%
\subsection*{3.14}
\begin{translation}[hp03_014]
Problems such as the great afflictions [and] death and so forth dissolve, and that is why the most wise call it the `great seal.'
\end{translation}
% MD: Now I prefer jīryante to hīyante. The latter is suppoted only α1 and α2, may be an outlier. As an explanation of the name mahāmudrā, √jṝ seems to make more sense. The kleśa etc. decay after being confined (mudrita) or so. done

@@

\begin{philcomm}[hp03_018]
Two readings of the second quarter are well-attested: \emph{jarāmṛtyuvināśinī} (\alphaOne, etc.) and \emph{mahāsiddhikarī nṛṇām} (\alphaThree, etc.). While the first reading is possible, the play on \emph{mahāsiddhi} and \emph{mahāmudrā} seems more likely original, as seen in the source text, the \emph{Vivekamārtaṇḍa} (without significant variants).\lb

%Manuscripts of three groups (\textalpha, \textbeta\ and \textgamma), which are important stemmatically, have the reading \emph{jarāmṛtyuvināśinī} (and \alphaTwo\ has \emph{nṛṇāṃ mṛtyuvināśinī}) for the second verse quarter. While this reading is possible, the play on \emph{mahāsiddhi} and \emph{mahāmudrā} seems more likely original, as seen in the source text, the \emph{Vivekamārtaṇḍa} (without significant variants), and the \texteta\ and \textzeta\ groups. \lb %  check the stemma tree -- α1+α2 vs α3

@@

\begin{philcomm}[hp03_020]
The referent of \emph{madhye} is uncertain. The verse is derived from the \emph{Amaraugha}, and the \emph{Amṛtasiddhi} makes no mention of a place to focus the mind in its treatment of \emph{mahābandha} (it does however instruct the yogi to place the mind at the \emph{catuṣpatha} in its teachings on \emph{mahāmudrā}). Bhavadevamiśra (7.196), Brahmānanda (3.20) and Bālakṛṣṇa (5.24) take it to mean the central channel. It could also plausibly mean the region between the chest and perineum, or perhaps the place between the eyebrows. At 3.24 this practice is said to make the mind reach Kedāra, which is sometimes located between the eyebrows (see Mallinson 2007: 214 n.285; Birch 2019: 967 n.57).
\end{philcomm}

%%%%%%%%%%
\subsection*{3.21}
\begin{translation}[hp03_021]
[The yogi] should hold the breath as long as possible and exhale slowly. And having practised it on the left side, he should practise it on the right side. 
\end{translation}
% MD: pāda a: adopt yathāśaktyā as in 2.7?
% pāda c: dakṣāṅge ca samabhyaset or dakṣāṅge punar abhyaset instead of dakṣiṇāṅge samabhyaset?

@@

\begin{philcomm}[hp03_021]
This section in the description of \emph{mahābandha} has been subject to various revisions in the manuscript groups. The shortest version occurs in \alphaOne\ and \alphaTwo, as well as the \textbeta\ group and the \emph{Haṭharatnāvalī}, which omit verses 3.21 and 3.22. This omission results in the somewhat odd juxtaposition of two lines stating that this is \emph{mahābandha}. \lb

@@

\begin{philcomm}[hp03_022]
This verse expresses an alternative to the application of the chin lock in \emph{mahā\-bandha} (mentioned in 3.20). It is found in all groups but is notably absent in \alphaOne\ and \alphaTwo, and the \textdelta\ group. At present we cannot be certain if this verse was written by Svātmārāma because it does not reflect textual teachings on \emph{mahābandha} in so far as no other text known to us advocates the use of the tongue rather than the chin lock in \emph{mahābandha}. If the verse was composed by Svātmārāma, it was omitted early in the transmission by someone who did not agree with the alternative teaching.\lb

This verse occurs in the \emph{Yogacintāmaṇi} in the middle of a quotation attributed to Īśvara. The other verses of the quoted passage are found in the \emph{Śivasaṃhitā} (4.37–42), but the verse in question is not reported in the critical edition of the \emph{Śivasaṃhitā} (2018). The verse is absent in another passage on \emph{mahābandha} that the author of the \emph{Yogacintāmaṇi} cites and attributes to the \emph{Haṭhapradīpikā}.\lb

@@

\begin{philcomm}[hp03_024]
The reading \emph{ūrdhvaṃgatibodhakaḥ} (`initiates an upward flow') is found in the source (\emph{Amaraugha} 25) and all \emph{Haṭhapradīpikā} witnesses except the \emph{Jyotsnā} (where the line is found earlier). While the \emph{Amaraughaprabodha} has the reading \emph{ūrdhvaṃgativiśodhanaḥ} (`purification of the upward flow'), the \emph{Amṛtasiddhi} (12.14) states that the chin-lock prevents the upward flow (\emph{ūrdhvaṃgatinirodhakaḥ}) in all the channels. The \textit{Amaraugha} is referring to the idea (likely accepted by Svātmārāma) that the root lock creates an upward flow in all the channels that prevents the elements and essences of the body from escaping (cf.~\emph{Amṛtasiddhi} 12.8–10).\lb

Triveṇī and Kedāra are pilgrimage sites, the former at Prayāga where the Gaṅgā and Yamunā meet, the latter in the Himālaya, near the source of the Gaṅgā. The bodily \emph{triveṇī} is located in the navel or heart by earlier Śaiva works (Birch 2019: 967). Here it may be the same as the \emph{trikūṭa} and located between the eyebrows (Mallinson 2007: 209 n.259). Brahmānanda, who does not identify a location for Triveṇī, understands Kedāra to be between the eyebrows. In the \emph{Khecarīvidyā} it is located on the back of the head above the nape of the neck (Mallinson 2007: 214 n.285). For other references on the location of Kedāra, see Birch 2019: 967 n.57.
\end{philcomm}

@@

\begin{philcomm}[hp03_027]
The term \emph{puṭadvaya} is referring back to the \emph{dvipuṭa} mentioned in verse 3.12. On the alchemical meaning and other interpretations by later commentators, see the note to 3.12.\lb

@@

\begin{philcomm}[hp03_028]
The \textalpha\ and \textgamma\ groups have °\emph{saṃbandhāj}, which is possible but somewhat awkward (i.e., `because of the union of the moon, sun and fire, the state of a dead person, which has arisen, leads to immortality'). The adopted reading °\emph{saṃbandho}, which is supported by \zetaTwo, \emph{Yuktabhavadeva} 7.200 and \emph{Jyotsnā} 3.28, makes better sense but its meaning is not as clear as the formulation in the source text (i.e., \emph{Amaraugha} 30).\lb

The compound \emph{mṛtāvasthā} (`the state of death') likely refers to a dead person in the sense that the yogi appears as if dead when the moon, sun and fire have united. In the version found in \emph{Jyotsnā} 3.28 and \etaOne, \etaTwo\ and \zetaThree, the final verse quarter has been rewritten to say that the yogi then exhales the breath (\emph{tato vāyuṃ virecayet}). This implies that \emph{mṛtāvasthā} is a breath retention (\emph{kumbhaka}), which is apparent in Brahmānanda's explanation:
\begin{quote}
The state of death that has arisen is the state of one who has died, [that is,] of one who is devoid of the breath because of the absence of movement of \emph{prāṇa} in the \emph{iḍā} and \emph{piṅgalā} channels. Then, immediately after that [state], [the yogi] exhales the breath, [that is,] he gradually releases it through the nostrils.
\end{quote}
\begin{quote}
\emph{mṛtasya prāṇaviyuktasyāvasthā mṛtāvasthā samutpannā bhavati, iḍāpiṅgalayoḥ prāṇasañcārābhāvāt} | \emph{tatas tadanantaraṃ vāyuṃ virecayen nāsikāpuṭābhyāṃ śanais tyajet} |
\end{quote}

@@

%%%%%%%%%%
\subsection*{3.29}
\begin{translation}[hp03_029]
Through practice, this great piercing bestows the great \emph{siddhi} [and] cures wrinkles, grey hair, and trembling. It is used by the best practitioners.
\end{translation}

@@

\begin{philcomm}[hp03_029]
In the third verse quarter, the reading \emph{°vedhaghnaḥ} (`cures wounds'?) in \textalpha\ and the other groups, the exception being \textgamma\ (\emph{°vegaghnaḥ}), is strange as it does not seem related to the other two symptoms of old age (i.e., wrinkles and grey hair) that this \emph{mudrā} can cure. It is perhaps a misreading of \emph{°vepaghnaḥ}, 
%which occurs in some manuscripts in the \textbeta\ (\getsiglum{J1}, \getsiglum{N12}), \textzeta\ (\getsiglum{J14}) and \textdelta\ (\getsiglum{J3}, \getsiglum{N16}, \getsiglum{N18}) groups, 
which occurs in some manuscripts in the \deltaOmega\ (\getsiglum{J1}, \getsiglum{J3}, \getsiglum{N16}, \getsiglum{N18}) and \textbeta\ (\getsiglum{N12} contaminated) and \textzeta\ (\getsiglum{J14} contaminated) groups, % by MD
as well as the \emph{Yogacintāmaṇi} and \emph{Jyotsnā}. The meaning of \emph{vepa}, `trembling,' is consistent with wrinkles and grey hair.

%The collocation \emph{valīpalitaroga} that we have adopted, found in \getsiglum{C8}, is quite common (e.g. \emph{Śāradātilaka} \emph{6.104, Ānandakanda} 1.7.13, .1.7.31).
%JB: I think we should adopt vepa on the strength of the marmasthāna entry (\item[valīpalitavepaghnaḥ] Ba5, J1, J3, J6, J14, N4a, N11, N12, N16, N18), and the fact that it's closer to vedha (alpha reading) than roga (which looks like an outlier in the marmasthāna data). 
% J1,J3,J6,N16,N18=6a, J14=4c, N11 =Gr2, N12 =4b,
% MD: These mss are not very strong. Is it not possible to interpret °vedhaghnaḥ as "destroys by piercing"? (like daṇḍaghna "striking with a stick" in the Manusmṛti) This verse only has no source in the middle of the long citation from Amaraugha, is probably authorial. Svātmārāma might have wanted to use the word vedha of mahāvedha, similar to dhauti/dhāvanti.
\end{philcomm}

@@

Various versions of this verse have been transmitted in \emph{āryā}, \emph{gīti} and \emph{anuṣṭubh} metres. We have adopted a version close to \alphaTwo, which has a slight metrical fault:
\begin{quote}
 chedanacālanadohaiḥ kalāṃ krameṇa pravardhayet tāvat |\\
 sā yāvad bhrūmadhyaṃ viśati tadānīṃ khecarīsiddhiḥ ||
\end{quote}
% alpha2
% chedanacālanaṃ dohaiḥ kalā kramaṇa pravardhaye tāvat
% sā yāvad bhrumadhyaṃ visati tadānī khecarīsiddhiḥ
The emendation of \emph{tadānīṃ} to \emph{tadā} renders the verse an \emph{āryā}. The word \emph{krameṇa} is well attested by manuscripts of the \textalpha, \textbeta, \texteta\ and  \textzeta\ groups, which all have unmetrical or corrupted versions.\lb 
% krameṇa is well attested
% close to alpha 2
% damaged in alpha 1: [damaged]ḥ kalāḥ krameṇa varddayet | sā yāva[d] bhṛmadhyaṃ viśa tadānīṃ khecarīsiddhiḥ || but it looks like someone has unsuccessfully tried to make it anuṣṭubh. Perhaps, suggests that a longer metre was original.

@@

%%%%%%%%%%
\subsection*{3.35}
\begin{translation}[hp03_035]
If the yogi turns back the tongue and remains [like that] for half an instant, he is instantly freed from disease, death, old age and the like.
\end{translation}

\begin{sources}[hp03_035]
\emph{Śivasaṃhitā} 3.91
\begin{versinnote}
\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
\tl{kṣaṇena mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_035]
\emph{Yogacintāmaṇi} f.~74r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{kalāṃ parāṅmukhīṃ kṛtvā tripathe parivartayet |\\+}
\tl{sā bhavet khecarī mudrā vyomacakraṃ tad ucyate ||\\+}
\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
\tl{viṣayair mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp03_035]
Although 3.35 is absent in \alphaOne, it is in \alphaTwo\ and \alphaThree, and also the \textgamma\ and \texteta\ groups. Other manuscripts have an additional line that gives \emph{vyomacakra} as an alternative name for \emph{khecarīmudrā}. This alternative name does not occur in any of the source texts known to have been used by Svātmārāma, but it may have been inspired by the name \emph{nabhomudrā}, which is what the \emph{Vivekamārtaṇḍa} calls \emph{khecarīmudrā} (\emph{Vivekamārtaṇḍa} 40).   
\end{philcomm}
